\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-12 article.tex
Finish explaining bug

The instructions included are
\lstinline!trunc! has only been added in order to help cover more generated LLVM. % FIXME
%with only four distinct block terminators 
%\begin{enumerate}
%    \item Unconditional \lstinline!br! followed by the label to branch to
%    \item Conditional \lstinline!br!  followed by an \lstinline!i1! condition and two labels
%    \item \lstinline!ret! followed by an optional return value
%    \item \lstinline!unreachable!
%\end{enumerate}
The branching for most of these is trivial: after successful execution the flow of all but \lstinline!br!, \lstinline!ret! and \lstinline!unreachable! will unconditionally attempt to execute the next instruction in memory (i.e. increment the instruction pointer by instruction length).

@@

\label{appendix:llvmis}
\begin{enumerate}
    \item binary operations \lstinline!add!, \lstinline!and!, \lstinline!ashr!, \lstinline!lshr!, \lstinline!mul!, \lstinline!or!, \lstinline!sdiv!,\\\lstinline!srem!, \lstinline!shl!, \lstinline!sub! and \lstinline!xor!
    \item integer comparison \lstinline!icmp! (conditions being \lstinline!eq!, \lstinline!ne!, \lstinline!sge!, \lstinline!sgt!, \lstinline!sle! and \lstinline!slt!)
    \item memory/address operations \lstinline!alloca!, \lstinline!gep!, \lstinline!load! and \lstinline!store!
    \item  \lstinline!mov! operations \lstinline!gep!, \lstinline!phi!, \lstinline!ptrtoint!, \lstinline!trunc!, and \lstinline!zext!
    \item control flow operations \lstinline!br!, \lstinline!call! and \lstinline!ret!
    \item a  nullary block terminator \lstinline!unreachable!
\end{enumerate}

@@

\begin{figure}[H]
   \begin{minipage}{0.58\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{llvm}
 ; ...
5:
 %6 = phi i32 [%7, %5], [1, %3]
 %7 = add i32 %6, 1
 %8 = mul i32 %7, %7
 %9 = icmp sgt i32 %8, %0
 br i1 %9, label %10, label %5
10:
 %11 = phi i32 [%0, %1], [0, %3], [%6, %5]
 ret i32 %11
}
     \end{minted}
     \caption{\texttt{dune exec build -{}- -t x86}}\label{fig:sqrt1.ll}
   \end{minipage}
   \begin{minipage}{0.40\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{gas}
# ...
# br i1 %9, label %10, label %5
movq	%r10, %rax
cmpq	$0, %rax
movq	%r8, %rax
movq	%rax, %rdi
je	_isqrt$5
movq	%rdi, %rax
movq	%rax, %r11
jmp	_isqrt$10
# ...
     \end{minted}
     \caption{TODO}\label{fig:sqrt1.s}
   \end{minipage}
\end{figure}
This was resolved by relying on conditional movs immediately prior to branching conditionally. Specifically, the \texttt{cmoveq} instruction was used, which only moves the value if the equals flag (i.e. \texttt{ZF}) is set.

>>>>

Some additional instructions are included, like \lstinline!trunc! and \lstinline!switch!, to widen the range of supported frontend languages with the clang compiler.
%been added in order to help cover more generated LLVM. % FIXME
%with only four distinct block terminators 
%\begin{enumerate}
%    \item Unconditional \lstinline!br! followed by the label to branch to
%    \item Conditional \lstinline!br!  followed by an \lstinline!i1! condition and two labels
%    \item \lstinline!ret! followed by an optional return value
%    \item \lstinline!unreachable!
%\end{enumerate}
%The branching for most of these is trivial: after successful execution the flow of all but \lstinline!br!, \lstinline!ret! and \lstinline!unreachable! will unconditionally attempt to execute the next instruction in memory (i.e. increment the instruction pointer by instruction length).

@@

%\label{appendix:llvmis}
%\begin{enumerate}
    %\item binary operations \lstinline!add!, \lstinline!and!, \lstinline!ashr!, \lstinline!lshr!, \lstinline!mul!, \lstinline!or!, \lstinline!sdiv!,\\\lstinline!srem!, \lstinline!shl!, \lstinline!sub! and \lstinline!xor!
    %\item integer comparison \lstinline!icmp! (conditions being \lstinline!eq!, \lstinline!ne!, \lstinline!sge!, \lstinline!sgt!, \lstinline!sle! and \lstinline!slt!)
    %\item memory/address operations \lstinline!alloca!, \lstinline!gep!, \lstinline!load! and \lstinline!store!
    %\item  \lstinline!mov! operations \lstinline!gep!, \lstinline!phi!, \lstinline!ptrtoint!, \lstinline!trunc!, and \lstinline!zext!
    %\item control flow operations \lstinline!br!, \lstinline!call! and \lstinline!ret!
    %\item a  nullary block terminator \lstinline!unreachable!
%\end{enumerate}

@@

\begin{figure}[H]
   \begin{minipage}{0.58\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{llvm}
 ; ...
5:
 %6 = phi i32 [%7, %5], [1, %3]
 %7 = add i32 %6, 1
 %8 = mul i32 %7, %7
 %9 = icmp sgt i32 %8, %0
 br i1 %9, label %10, label %5
10:
 %11 = phi i32 [%0, %1], [0, %3], [%6, %5]
 ret i32 %11
}
     \end{minted}
     \caption{\texttt{dune exec build -{}- -t x86}}\label{fig:sqrt1.ll}
   \end{minipage}
   \begin{minipage}{0.40\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{gas}
# ...
# br i1 %9, label %10, label %5
movq	%r10, %rax
cmpq	$0, %rax
movq	%r8, %rax
movq	%rax, %rdi
je	_isqrt$5
movq	%rdi, %rax
movq	%rax, %r11
jmp	_isqrt$10
# ...
     \end{minted}
     \caption{TODO}\label{fig:sqrt1.s}
   \end{minipage}
\end{figure}

\noindent This was resolved by relying on conditional \texttt{mov}s immediately prior to the conditional branch (i.e. \texttt{je}). Specifically, the \texttt{cmoveq} instruction was used, which only moves from source to destination if the equality flag (i.e. \texttt{ZF}) is set.
Unfortunately the \texttt{cmoveq} instruction doesn't support indirect addressing, so in order to support spilled variables both the values need to be moved into registers as can be seen in Figure \ref{fig:fixed-ind-sqrt1.s}.

\begin{figure}[H]
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{gas}
# ...
# br i1 %9, label %10, label %5
movq	%rdx, %rax
cmpq	$0, %rax
movq	%rbx, %rax
cmoveq	%rax, %rsi
je	_isqrt$5
movq	%rsi, %rax
movq	%rax, %rdx
jmp	_isqrt$10
# ...
     \end{minted}
     \caption{TODO}\label{fig:fixed-sqrt1.s}
   \end{minipage}
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{gas}
# ...
# br i1 %13, label %14, label %9
movq	-88(%rbp), %rax
cmpq	$0, %rax
movq	-72(%rbp), %rax
movq	-16(%rbp), %rcx
cmoveq	%rax, %rcx
movq	%rcx, -16(%rbp)
je	_main$9
movq	-16(%rbp), %rax
# ...
     \end{minted}
     \caption{\texttt{-a greedy -n 0}}\label{fig:fixed-ind-sqrt1.s}
   \end{minipage}
\end{figure}
